Neighbor-Based Decentralized Training Strategies for Multi-Agent Reinforcement Learning

### 2024-10-12 paper-2024-sac-marl-neighbouring-strategies.tex
chore: minor on the experimental setup

\subsection{State of the Art}\label{sec:sota}
MARL is a longstanding research area with a rich literature, starting 
from seminal works like~\cite{tan1993multi} and~\cite{busoniu2008comprehensive} to more recent contributions like~\cite{gronauer2022multi} and~\cite{canese2021multi}.
%
In the following, we briefly discuss some of the most relevant approaches in the field, which lead to the motivation for our work.
\subsubsection{Independent Learners}
In independent learners, each agent learns its policy independently, disregarding the actions and policies of other agents. 
This straightforward approach, among the earliest explored in MARL, often relies on variants of Q-learning. 
%
Several algorithms have attempted to address these coordination challenges within the independent learner paradigm. 
Basic decentralized Q-learning~\cite{tan1993multi} allows each agent to learn its Q-values independently but often converges to suboptimal joint policies due to a lack of coordination. 
Subsequent work introduced heuristic-based extensions to improve coordination.
Approaches like distributed Q-learning~\cite{lauer2000algorithm} incorporate optimism in Q-value updates and equilibrium selection mechanisms to mitigate the convergence to shadowed equilibria (i.e., equilibria that are unstable due to the independent learning processes). 
Instead, algorithms such as hysteretic Q-learning~\cite{matignon2007hysteretic}, lenient Q-Learning~\cite{bloembergen2010lenient} and win-or-learn fast policy hill climbing~\cite{bowling2002multiagent} employ adaptive learning rates to address the issue of alternating exploration between agents, improving robustness and convergence.
%
Empirical comparisons of these algorithms highlight the trade-offs in performance across various MARL tasks and coordination challenges~\cite{matignon2012independent}. 
While initially developed for tabular settings, some of these methods, like hysteretic Q-learning~\cite{palmer2017lenient} and lenient Q-learning, have been extended to deep reinforcement learning settings. 
However, these independent approaches, even with deep learning, often exhibit instability and slow convergence in complex environments and fundamentally lack explicit mechanisms for inter-agent coordination. 
Consequently, more sophisticated MARL algorithms have been developed that explicitly account for the interactions and policies of other agents to facilitate more effective coordination.
\subsubsection{Neighboring-Based Methods}
Neighboring-based methods in MARL address the scalability and communication overhead challenges of fully centralized approaches by restricting agent interactions to a local neighborhood. 
This introduces a structure where agents communicate and coordinate with a limited set of neighbors, reducing the complexity of joint action and policy learning.
%
One prominent category within neighboring-based methods is \emph{consensus-based} algorithms. 
These algorithms aim to reach a consensus among agents on shared values or policies through local information exchange. 
Early work like QD-learning~\cite{kar2012qd} extends Q-learning with a consensus update term, 
enabling agents to converge towards optimal Q-values despite only receiving local rewards.
%  
Recent advances in consensus-based MARL leverage deep learning and actor-critic architectures, 
combining local policy updates with consensus updates on critic parameters or gradients, 
as seen in the recent works~\cite{7434032,pennesi2010distributed,zhang2018fully,DBLP:journals/jmlr/CiosekW20}.

@@

\section{Neighboring-Based Distributed Learning Strategies}\label{sec:neighboring}
This section introduces three neighboring-based distributed learning strategies for multi-agent reinforcement learning.
These strategies leverage local interactions between agents to improve learning efficiency and coordination.
For this investigation, we focus on deep Q-learning as the underlying learning algorithm,
and we consider the following strategies: k-nearest neighbor averaging, k-nearest neighbor consensus, and experience sharing.

\paragraph{K-Nearest Neighbor}
Since these strategies are in general based on the notion of k-nearest neighbors,
we first introduce this concept and then describe each strategy in detail.
Formally, consider a \emph{weighted} communication graph $\mathcal{G} = (N, E, W)$, where $N$ is the set of agents, where $W: E \rightarrow \mathbb{R}^+$ is a weight function assigning a positive weight to each edge.  
Let $d(i,j)$ denote the shortest path length between agents $i$ and $j$ in the weighted graph $\mathcal{G}$. 
This can be calculated, for instance, using Dijkstra's algorithm. 
The length of a path is the sum of the weights of the edges along the path. 
If no path exists between $i$ and $j$, we set $d(i,j) = \infty$.
%
Then, the set of $k$-nearest neighbors of agent $i$ is defined as:
$$ \mathcal{N}_k(i) = { j \in N \setminus {i} \mid \text{rank}(d(i,j)) \leq k } $$
where $\text{rank}(d(i,j))$ denotes the rank of the distance $d(i,j)$ among all distances from agent $i$ to other agents in $N \setminus {i}$, sorted in ascending order (e..g, $\text{rank}(d(i,j)) = 1$ for the smallest distance).
Ties can be broken arbitrarily.

\paragraph{K-Nearest Neighbor Averaging}
This approach averages an agent's Q-values with those of its nearest neighbors. 
Each agent maintains a local Q-network and shares its Q-values with its neighbors. 
The neighbors average the received Q-values and use the result to update their own Q-networks. 
This leverages the collective knowledge of neighboring agents to improve learning.
%
Formally, giving $\theta^t_i, i \in N$ the parameters of the Q-networks of the agent $i$ at the time $t$,
the update rule for agent $i$ at time $t+1$ is:
\begin{equation}
  \theta^{t+1}_i = \theta^t_i + \alpha \sum_{j \in \mathcal{N}_k(i)} \frac{1}{k} \theta^t_j
\end{equation}

@@

To validate our approach, we developed a custom learning scenario using three State-of-the-Art frameworks 
  in Reinforcement Learning and Deep Learning, namely: \emph{Ray}~\cite{DBLP:conf/osdi/MoritzNWTLLEYPJ18} 
  (in particular, its module \emph{RLlib}~\cite{DBLP:conf/icml/LiangLNMFGGJS18}),
  \emph{Gymnasium}~\cite{DBLP:journals/corr/abs-2407-17032}
  and \emph{PyTorch}~\cite{DBLP:conf/nips/PaszkeGMLBCKLGA19}.

We ran a total of $35$ training experiments, namely: $5$ different methods 
  (centralized, distributed and neighbor based approaches), and for each method we ran multiple experiments
  varying the random seed (7 seeds) to obtain stronger results. 
%
Each experiment consisted of $50$ training episodes.

>>>>

\subsection{State of the Art}\label{sec:sota}
MARL is a longstanding research area with a rich literature, starting 
from seminal works like~\cite{tan1993multi} and~\cite{busoniu2008comprehensive} to more recent contributions like~\cite{gronauer2022multi} and~\cite{canese2021multi}.
%
In the following, we briefly discuss some of the most relevant approaches in the field, which lead to the motivation for our work.
\subsubsection{Independent Learners}
In independent learners, each agent learns its policy independently, disregarding the actions and policies of other agents. 
This straightforward approach, among the earliest explored in MARL, often relies on variants of Q-learning. 
%
Several algorithms have attempted to address these coordination challenges within the independent learner paradigm. 
Basic decentralized Q-learning~\cite{tan1993multi} allows each agent to learn its Q-values independently but often converges to suboptimal joint policies due to a lack of coordination. 
Subsequent work introduced heuristic-based extensions to improve coordination.
Approaches like distributed Q-learning~\cite{lauer2000algorithm} incorporate optimism in Q-value updates and equilibrium selection mechanisms to mitigate the convergence to shadowed equilibria (i.e., equilibria that are unstable due to the independent learning processes). 
Instead, algorithms such as hysteretic Q-learning~\cite{matignon2007hysteretic}, lenient Q-Learning~\cite{bloembergen2010lenient} and win-or-learn fast policy hill climbing~\cite{bowling2002multiagent} employ adaptive learning rates to address the issue of alternating exploration between agents, improving robustness and convergence.
%
Empirical comparisons of these algorithms highlight the trade-offs in performance across various MARL tasks and coordination challenges~\cite{matignon2012independent}. 
While initially developed for tabular settings, some of these methods, like hysteretic Q-learning~\cite{palmer2017lenient} and lenient Q-learning, have been extended to deep reinforcement learning settings. 
However, these independent approaches, even with deep learning, often exhibit instability and slow convergence in complex environments and fundamentally lack explicit mechanisms for inter-agent coordination. 
%Consequently, more sophisticated MARL algorithms have been developed that explicitly account for the interactions and policies of other agents to facilitate more effective coordination.
\subsubsection{Neighboring-Based Methods}
Neighboring-based methods in MARL address the scalability and communication overhead challenges of fully centralized approaches by restricting agent interactions to a local neighborhood. 
This introduces agents communication and coordination with a limited set of neighbors, reducing the complexity of full joint action and policy learning.
%
One prominent category within neighboring-based methods is \emph{consensus-based} algorithms. 
These algorithms aim to reach a consensus among agents on shared values or policies through local information exchange. 
Early work like QD-learning~\cite{kar2012qd} extends Q-learning with a consensus update term, 
enabling agents to converge towards optimal Q-values despite only receiving local rewards.
%  
Recent advances in consensus-based MARL leverage deep learning and actor-critic architectures, 
combining local policy updates with consensus updates on critic parameters or gradients, 
as seen in the recent works~\cite{7434032,pennesi2010distributed,zhang2018fully,DBLP:journals/jmlr/CiosekW20}.

@@

\section{Neighboring-Based Distributed Learning Strategies}\label{sec:neighboring}
This section introduces three neighboring-based distributed learning strategies for multi-agent reinforcement learning.
These strategies leverage local interactions between agents to improve learning efficiency and coordination.
For this investigation, we focus on deep Q-learning as the underlying learning algorithm,
and we consider the following strategies: k-nearest neighbor averaging, k-nearest neighbor consensus, and experience sharing.

\paragraph{K-Nearest Neighbor Definition}
The proposed strategies rely on the concept of k-nearest neighbors.
Formally, consider a \emph{weighted} communication graph $\mathcal{G} = (N, E, W)$, where $N$ is the set of agents, where $W: E \rightarrow \mathbb{R}^+$ is a weight function assigning a positive weight to each edge.  
Let $d(i,j)$ denote the shortest path length between agents $i$ and $j$ in the weighted graph $\mathcal{G}$. 
This can be calculated, for instance, using Dijkstra's algorithm. 
The length of a path is the sum of the weights of the edges along the path. 
If no path exists between $i$ and $j$, we set $d(i,j) = \infty$.
%
Then, the set of $k$-nearest neighbors of agent $i$ is defined as:
$$ \mathcal{N}_k(i) = { j \in N \setminus {i} \mid \text{rank}(d(i,j)) \leq k } $$
where $\text{rank}(d(i,j))$ denotes the rank of the distance $d(i,j)$ among all distances from agent $i$ to other agents in $N \setminus {i}$, sorted in ascending order (e..g, $\text{rank}(d(i,j)) = 1$ for the smallest distance).
Ties can be broken arbitrarily.

\paragraph{K-Nearest Neighbor Averaging}
\sloppy
This approach averages an agent's Q-values with those of its nearest neighbors. 
Each agent maintains a local Q-network and shares its Q-values with its neighbors. 
The neighbors average the received Q-values and use the result to update their own Q-networks. 
This leverages the collective knowledge of neighboring agents to improve learning.
%
Formally, giving $\theta^t_i, i \in N$ the parameters of the Q-networks of the agent $i$ at the time $t$,
the update rule for agent $i$ at time $t+1$ is:
\begin{equation}
  \theta^{t+1}_i = \theta^t_i + \alpha \sum_{j \in \mathcal{N}_k(i)} \frac{1}{k} \theta^t_j
\end{equation}

@@

To validate our approach, we developed a custom learning scenario using three state-of-the-art frameworks 
  in RL and DL, namely: \emph{Ray}~\cite{DBLP:conf/osdi/MoritzNWTLLEYPJ18} 
  (in particular, its module \emph{RLlib}~\cite{DBLP:conf/icml/LiangLNMFGGJS18}),
  \emph{Gymnasium}~\cite{DBLP:journals/corr/abs-2407-17032}
  and \emph{PyTorch}~\cite{DBLP:conf/nips/PaszkeGMLBCKLGA19}.

We ran a total of $35$ training experiments, 
namely: $5$ different methods %% TODO check how many ex are discussed
  (centralized, distributed and the three neighbor strategies discussed).
Each method was evaluated across 10 distinct random seeds to enhance the statistical robustness of the observed results.
%
Each experiment consisted of $50$ training episodes.
